---
title: "CLaRe: A General Evaluation Framework for Selecting Compact near-lossless Latent Representations of High-Dimensional Object Data"
output: github_document
---

This repository contains the analysis scripts and LaTeX files accompanying the __*revised*__ manuscript:
"CLaRe: A General Evaluation Framework for Selecting Compact near-lossless Latent Representations of High-Dimensional Object Data"* by Emma Zohner, Edward Gunning, Giles Hooker and Jeffrey Morris.

The repository provides:
  
1. Scripts for reproducing analyses presented in the manuscript.
  
 <!-- 2. LaTeX files for generating the manuscript, including figures, tables, and supplementary materials. -->
  
2. Integration with the [**GLaRe** R package](https://github.com/edwardgunning/GLaRe), which implements the CLaRe framework.

---

## Installing the GLaRe Package

This manuscript relies heavily on the [**GLaRe** R package](https://github.com/edwardgunning/GLaRe).
You should run the following commands to install the latest version from GitHub:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github(repo = "https://github.com/edwardgunning/GLaRe")
library(GLaRe)
```

## Python Environment

The autoencoder representation uses a `keras` implementation. This framework and code was developed using the original [`keras`](https://cran.r-project.org/web/packages/keras/index.html) R package, which is now deprecated. To ensure reproducibility for this manuscript, we have created the following virtual Python environment (`glare-legacy`):

```{r}
library(reticulate)
use_python("~/.virtualenvs/glare-legacy/bin/python", required = TRUE)
py_config()
```

To create a suitable virtual environment on your machine, you should run:



---

## Repository Overview

### Contents
- [`code/`](code/): R scripts for reproducing analyses and figures.
- [`main/`](main/): LaTeX files for individual parts of the manuscript, including main text, appendices, and references.
- [`data/`](data/): Datasets used in the analyses and generated results.
- [`figures/`](figures/): Generate figures for analysis.
- [`README.md`](README.md): This file, providing an overview of the repository.
- [`computation-time-results.md`](computation-time-results.md): Computation times for the analysis.

## Getting Started

### Prerequisites

- R (version >= 4.0.0)
- RStudio (optional)
- LaTeX distribution (e.g., TeX Live, MikTeX, Overleaf for online compilation)
- The following R packages:
  - [**GLaRe** (available via GitHub)](https://github.com/edwardgunning/GLaRe)
  - Additional dependencies: `ggplot2`, `dplyr`, `tidyr`, `scales`, `gridExtra`.
- __IMPORTANT__: Python Environment

---
