---
title: "CLaRe: A General Evaluation Framework for Selecting Compact near-lossless Latent Representations of High-Dimensional Object Data"
output: github_document
---

This repository contains the analysis scripts and LaTeX files accompanying the __*revised*__ manuscript:
<center>
**"CLaRe: A General Evaluation Framework for Selecting Compact near-lossless Latent Representations of High-Dimensional Object Data"**
</center>
by *Emma Zohner, Edward Gunning, Giles Hooker* and *Jeffrey Morris*.

The repository provides:
  
1. Scripts for reproducing analyses presented in the manuscript.

2. Integration with the [**GLaRe** R package](https://github.com/edwardgunning/GLaRe), which implements the CLaRe framework.

---

## Installing the `GLaRe` Package

This manuscript relies heavily on the [**GLaRe** R package](https://github.com/edwardgunning/GLaRe).
You should run the following commands to install the latest version from GitHub:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github(repo = "https://github.com/edwardgunning/GLaRe")
library(GLaRe)
```

## Python Environment (Important)

The autoencoder representation uses a `keras` implementation. This framework and code was developed using the original [`keras`](https://cran.r-project.org/web/packages/keras/index.html) R package, which is now deprecated. To ensure reproducibility for this manuscript, we have created the following virtual Python environment (`glare-legacy`):

```{r}
library(reticulate)
use_python("~/.virtualenvs/glare-legacy/bin/python", required = TRUE)
py_config()
```

To create a suitable virtual environment on your machine, you should run:

```bash
python3.9 -m venv .venv
source .venv/bin/activate
pip install --upgrade pip
pip install -r requirements-python.txt
```

[requirements-python.txt located here](requirements-python.txt).

Once the environment has been created, the Python environment can be loaded from R using:

```{r, eval=FALSE}
library(reticulate)
use_python(".venv/bin/python", required = TRUE)
py_config()
```

or convenience, the analysis scripts source the file [`load_Python_legacy_env.R`](load_Python_legacy_env.R), which selects the project-local `.venv` environment when available. 

On the author's machine, this falls back to the original local environment:

```{r, eval=F}
library(reticulate)

project_python <- file.path(getwd(), ".venv", "bin", "python")

if (file.exists(project_python)) {
  reticulate::use_python(project_python, required = TRUE)
} else {
  reticulate::use_python("~/.virtualenvs/glare-legacy/bin/python", required = TRUE)
}

print(reticulate::py_config())
```


---

## Repository Overview

### Contents
- [`code/`](code/): R scripts for reproducing analyses and figures. The running order is as follows:
  - **Information Loss Figure 1** [01-information-loss-figure.R](code/01-information-loss-figure.R) creates the example for Figure 1 of the manuscript.
  - **Distributional Summaries Figure 2**: [02-generror-distribution-summaries-figure.R](code/02-generror-distribution-summaries-figure.R) creates the Figure 2 in the paper, containing the three different plots of the generalization error distribution (GLaRe plot, dotplot, heatmap). The example here is PCA on the `phoneme` data.
  - **Datasets Figure 3**: [03-data-objects-plot.R](code/03-data-objects-plot.R) plots the datasets for the case studies (Figure 3 in the paper).
  - **Eye/ Glaucoma Case Study**: [04.1-run-eye-analysis.R](code/04.1-run-eye-analysis.R),[04.2-plot-eye-results.R](code/04.2-plot-eye-results.R), [04.3-eye-reconstruction-plus-other.R](code/04.3-eye-reconstruction-plus-other.R) deal with the running and plotting of the analysis of the eye data in the case study part 1.
  - **Gels Case Study**: [05.1-run-gels-pca-dwt.R](code/05.1-run-gels-pca-dwt.R), [05.2-run-gel-analysis-ae-batch.R](code/05.2-run-gel-analysis-ae-batch.R),  [05.3-run-gel-analysis-ae-worker.R](code/05.3-run-gel-analysis-ae-worker.R), [05.4-combine-gels-ae-results.R](code/05.4-combine-gels-ae-results.R), [code/05.5-plot-gels-results.R](code/05.5-plot-gels-results.R), [05.6-plot-gels-reconstruction.R](code/05.6-plot-gels-reconstruction.R) deal with the running and plotting of the analysis of the gels data in the case study part 2. Note that because of memory limitations on the author's laptop, we had to call GlaRe in a loop for each candidate qualifying dimension separately. The worker script being called is [05.3-run-gel-analysis-ae-worker.R](code/05.3-run-gel-analysis-ae-worker.R) and it is being called within a loop by [05.2-run-gel-analysis-ae-batch.R](code/05.2-run-gel-analysis-ae-batch.R).
  - **MNIST Case Study**: [06.1-run-mnist-analysis.R](code/06.1-run-mnist-analysis.R), [06.2-plot-mnist-results.R](code/06.2-plot-mnist-results.R), [06.3-mnist-reconstruction.R](code/06.3-mnist-reconstruction.R) contain the scripts for the mnist case study part 3.
  - **Sample Size Experiments Case Study** [07.1-sample-size-experiment-seed-01.R](code/07.1-sample-size-experiment-seed-01.R), [](code/07.2-sample-size-experiment-seed-02.R), [07.3-sample-size-experiment-seed-03.R](code/07.3-sample-size-experiment-seed-03.R).
  - **Additional (Appendix) Case Study `phoneme`** [08-phoneme-data.R](code/08-phoneme-data.R).
  - **Additional Analysis (Appendix) Multivariate Functional Data**: [additional-multivariate-functional-data.R](code/additional-multivariate-functional-data.R). Requires `fda` R package.
  - **Additional Data Example using Quantile Functions**: [additional-revision-quantiles.R)](code/additional-revision-quantiles.R) which uses the following helper functions in the file [additional-helpers-for-quantiles.R](code/additional-helpers-for-quantiles.R).
  - **Settings for `ggplot2` themes**: [theme_gunning.R](code/theme_gunning.R)
- [`data/`](data/): Datasets used in the analyses and generated results.
  - **DATASETS**:
    - The cleaned Glaucoma data is stored in [Y_outlier_removed.txt](data/Y_outlier_removed.txt).
    - The gels data is stored in [proteomic_gels.RData](data/proteomic_gels.RData)
    - We simply load the MNIST data using the `keras` package, running: `mnist <- keras::dataset_mnist()`.
    - We load the `phoneme` dataset directly from its website:
    ```{r eval = FALSE}
    PH_path <- "https://www.math.univ-toulouse.fr/~ferraty/SOFTWARES/NPFDA/npfda-phoneme.dat"
    PH <- readr::read_table(file = PH_path, col_names = FALSE)
    ```
    - Instructions for downloading the GaitRec dataset directly from its link are contained in [additional-multivariate-functional-data.R](code/additional-multivariate-functional-data.R).
- [`figures/`](figures/): Generate figures for analysis.
- [`README.md`](README.md): This file, providing an overview of the repository.
- [`computation-time-results.md`](computation-time-results.md): Computation times for the analysis.



## Additional Reproducibility Details

At the end of the analysis workflow, the computational environment can be recorded using:

```{r}
sessionInfo()

library(reticulate)
source("load_Python_legacy_env.R")
py_config()

py_run_string("
import sys
import numpy
import tensorflow as tf
import keras

print('Python executable:', sys.executable)
print('Python version:', sys.version)
print('NumPy:', numpy.__version__)
print('TensorFlow:', tf.__version__)
print('Keras:', keras.__version__)
")
```

<!-- For the author's final run, this information was saved using: -->

<!-- ```bash -->
<!-- Rscript code/99-session-info.R > session-info.txt -->
<!-- ``` -->

<!-- The file `session-info.txt` records the R session, loaded package versions, Python executable, Python package versions, system information, and Git commit used for the final reproducibility check. -->


